\vspec{} and Friends: A suite of utilities to model spectroscopic phase curves of 3D exoplanet atmospheres in the presence of stellar variability.

### 2024-07-24 src/tex/ms.tex
Rewrite GCM section

\section{Simplified Climate Model}
\vspec{} allows the user to upload any desired GCM. However, for questions that focus on detectability,
the specific physics of the GCM are less important than properties like day/night temperature, surface pressure, or
the major absorbers in the atmosphere. To simplify these studies, \vspec{} has a built-in GCM that takes in just a
few parameters and produces a simple-yet-robust exoplanet climate. Simplicity allows the user to know everything about
the model and reduce surprises, but its basis in thermodynamics and robust energy balance make this model fit for publication-level science \TJ{is this really true? probably not}.

The \vspec{} GCM centers around a 2D temperature map of the planet's surface. Based on \citet{cowan2011}, this map
is constructed completely given the incident flux from the star (parameterized by the host \teff and radius
and the planet's semimajor axis and Bond albedo) and the unitless thermal redistribution efficiency $\epsilon$.
This efficiency is 0 for a planet that reradiates instantly and $\gg 1$ for an atmosphere that efficiently redistributes heat to its night side.
Figure \ref{fig:thermal_inertia} shows a temperature map example for an Earth-like planet with $\epsilon = 2\pi$.
These temperature maps are validated by the energy balance of the planet; \vspec{} computes the incident and emergent
flux across the planet and asserts that they agree.

@@

Now that we have the surface temperature at each GCM coordinate we can build the atmosphere on top of it.
We choose a surface pressure for our model and generate a pressure profile as a function of layer. It is
important to note that the choice of the pressure profile -- provided it is scaled to the surface pressure and sampled in a reasonable manner -- is
arbitrary and will not affect the spectrum of the planet. This is because PSG calculates the altitude and thickness of each layer based on the
pressure and surface gravity; the pressure profile is essentially a proxy for the sampling of atmosphere layers as a function of altitude.

Next we create an adiabatic temperature profile at each GCM coordinate. An adiabat assumes that
\begin{equation}
    T^\gamma P^{1 - \gamma} = \text{const.}
\end{equation}
where $\gamma$ is the adiabatic index. Therefore the temperature profile can be computed
\begin{equation}
    T = T_{\rm surf} \left(\frac{P}{P_{\rm surf}} \right)^{1 - \frac{1}{\gamma} }
\end{equation}

@@

The caveats of this approach lie in the simplified physics of the atmosphere structure. The heat-redistribution model does not
capture the effects of latitudinal mixing; an adiabatic temperature profile will never exhibit the temperature inversion
that we know form in complex atmospheres like Earth's. Similarly, setting a constant abundance does not allow for water vapor to
vary with temperature as it does on Earth. We must also be careful not to add so much of an absorbing species that the energy balance of
the atmosphere would change; instead we treat them as trace species in an otherwise transparent atmosphere. Our simplified approach
allows us to easily examine a large parameter space of atmospheres whose behaviors are easy to predict.

>>>>

\section{Simplified Climate Model}
\vspec{} allows users to run models with any climate model they desire (so long as it can be converted into the \texttt{pypsg.PyGCM} format) and even comes with utilities to work with PSG/GlobES binary files, ExoCAM \citep{wolf2022}, the Whole Atmosphere Community Climate Model \citep[WACCM,][]{marsh2013}, and ExoPlaSim \citep{paradise2022} without any custom code from the user.

However, as a full global circulation model can take days to months to converge, there are many cases where a ``quick and dirty'' 3D climate model may be sufficient. Built into \vspec{} is such a ``faux'' climate model -- one that does not contain any of the circulation physics of the above sophisticated codes, but instead constructs a 3D planetary atmosphere in a fraction of a second from just a few parameters. Such a model can be advantageous, even, because it is extremely predictable; a feature in the phase curve can be traced back to the model parameters that cause it.

The \vspec{} GCM centers around a 2D temperature map of the planet's surface. Based on \citet{cowan2011}, this map
is constructed completely given the incident flux from the star (parameterized by the host \teff~and radius
and the planet's semimajor axis and Bond albedo) and the unitless thermal redistribution efficiency $\epsilon$.
This efficiency is 0 for a planet that reradiates instantly and $\gg 1$ for an atmosphere that efficiently redistributes heat to its night side.
Figure \ref{fig:thermal_inertia} shows a temperature map example for an Earth-like planet with $\epsilon = 2\pi$.
These temperature maps are validated by the energy balance of the planet; \vspec{} computes the incident and emergent
flux across the planet and asserts that they agree.

@@

To compute the surface temperature we first compute the temperature at the equator as a function of longitude. This is done by integrating Equation (10) of \citet{cowan2011}
\begin{equation}
    \frac{d \tilde{T}}{d \Phi} = \frac{1}{\epsilon}(\max{(\cos{\Phi}, 0)} - \tilde{T}^4)
\end{equation}
Where $\tilde{T}$ is the dimensionless temperature defined as $T/T_0$, or the temperature normalized by a local fiducial value. $\Phi$ is the angle from the substellar point, so we integrate if from $-\pi$ to $\pi$. This integral is done numerically in Figure \ref{fig:cowan_curves}, the optimal scheme depending on the value of $\epsilon$. For $0 < \epsilon < 1$ we adopt a reflecting initial value problem (IVP) scheme, where we set $\tilde{T}=1$ at $\Phi=\pi$, integrate to $-\pi$, then use the last value as the initial value for an integration from $-\pi$ to $\pi$. Note that for extremely small values of $\epsilon$ (e.g. $10^{-6}$) the integration is slow, and will fail if $\epsilon=0$; we recommended a lower limit of $\epsilon=10^{-4}$. When $1\le \epsilon < 10$ we solve the integral using a boundary value problem (BVP) scheme. For each of these two methods we employ the SciPy implementation (\texttt{solve\_ivp()} and \texttt{solve\_bvp()}, respectively). When $\epsilon \ge 10$ it is more efficient to use the analytic approximation given by \citet{cowan2011} Equation (A2).

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/cowan_fig1.pdf}
    \caption{
        Reproduction of \citet{cowan2011} Figure 1 using the methods described in the text. These curves show the dimensionless temperature at the equator of a planet as a function of angle from substellar point $\Phi$ for thermal inertia $\epsilon$. Note that the $\epsilon=0$ curve in the original figure is replaced with $\epsilon=10^{-4}$ (see text), and we do not include additional curves for mid-latitudes as in the original figure. As $\epsilon$ increases heat is moved from the dayside onto the nightside, and the temperature maximum shifts away from the substellar point.
    }
    \script{cowan_fig.py}
    \label{fig:cowan_curves}
\end{figure}

\subsection{Latitudinal Mixing}
The treatment by \citet{cowan2011} includes no mixing in the latitudinal direction, and the temperature tends to 0 as you approach the poles. We add a parameter $\alpha$ that is defined by the ratio between the polar temperature and the equitorial temperature at any given longitude. Let $f(\lambda;\alpha)$ be the temperature as a function of latitude $\lambda$ at an arbitrary longitude. If the surface is in equilibrium, the energy balance requires the outgoing energy flux to be independent of $\alpha$:
\begin{equation}
    \label{eq:e_out}
    E_{\rm out} = \sigma \int_0^{\pi/2} f(\lambda; \alpha)^4 \cos{\lambda}\,d\lambda
\end{equation}
Where $E_{\rm out}$ is the outbound energy flux of an infintesimal sliver of differential longitude.
In the case of $\alpha=0$ (i.e. no mixing) we can treat each latitude value independently, so the equilibrium temperature is
\begin{equation}
    f(\lambda; 0) = f(0;0) \cos^{1/4}{\lambda}
\end{equation}

Where $f(0;0)$ is the previously computed temperature at the equator (see Figure \ref{fig:cowan_curves}). Plugging this in to Equation (\ref{eq:e_out}) we see
\begin{equation}
    \label{eq:nomix}
    \begin{array}{cl}
        E_{\rm out} & = \sigma \int_0^{\pi/2} (f(0;0) \cos^{1/4}{\lambda})^4\cos{\lambda}\,d\lambda \\
        & = \sigma f^4(0;0) \int_0^{\pi/2} \cos^{2}{\lambda}\,d\lambda \\
        & = \frac{\pi}{4}\sigma f^4(0;0)
    \end{array}
\end{equation}

If instead we let $f(\lambda;1)$ be a constant function where the equator and pole have equal temperatures, then
\begin{equation}
    \label{eq:allmix}
    \begin{array}{cl}
        E_{\rm out} & = \sigma \int_0^{\pi/2} f(\lambda;1)^4 \cos{\lambda}\,d\lambda \\
        & = \sigma f^4(0;1) \int_0^{\pi/2} \cos{\lambda}\,d\lambda \\
        & = \sigma f^4(0;1)
    \end{array}
\end{equation}

We calculate $f(0,1)$ by setting Equations (\ref{eq:nomix} \& \ref{eq:allmix}) equal to find
\begin{equation}
    f(0;1) = \left(\frac{\pi}{4}\right)^{1/4} f(0;0) \approx 0.94 \,f(0;0)
\end{equation}

We can now write a general parameterization for the temperature as a function of latitude
\begin{equation}
    \label{eq:temp_gen}
    f(\lambda;\alpha) = f(0;0)\,\left(\alpha \frac{\pi}{4} + (1-\alpha) \cos{\lambda}\right)^{1/4}
\end{equation}

\subsection{Pressure and Temperature}
Combining Equation (\ref{eq:temp_gen}) with the numerically-computed equatorial temperatures yields a 2D temperature map that covers the surface of the planet. We construct a 3D atmosphere on top of this surface assuming the atmosphere is adiabatic with index $\gamma$. The user supplies both the adiabatic index and the surface pressure $P_{\rm surf}$. The temperature at some altitude $z$ is given by
\begin{equation}
    T_z = T_{\rm surf}\,\left( \frac{P_z}{P_{\rm surf}} \right)^{1-\frac{1}{\gamma}}
\end{equation}

@@
